Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-08-16 implementation/implementation.tex
New content

This subsection aims to preliminarily explore options in ABC SMC implementations. From the above results, we adopted multivariate normal kernel as the kernel for our real inference on observed data; Local M-NN kernels were also planned to be tested. In the real inference, 25:75 factors were also tried to see if it could help in recovering more features and resulting a better fit. because the improvement by using less data was inapparent, we did not try to change the target summary data size. Prior intervals were explored explicitly and gradually shrunk for our models.

@@

[separated ABC run on model 1, 2 and 3]

@@

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Mean value & Inter-quartile interval \\[0.5ex]
        \hline\hline
        $\lambda_N$          & 13.5       & [12.9, 14.2]            \\
        $a$                  & 0.0175     & [0.0145, 0.0207]        \\
        $\kappa_{N\beta}$    & 0.176      & [0.109, 0.234]          \\
        $\mu_N$              & 0.0239     & [0.0113, 0.0332]        \\
        $\nu_{N\Phi}$        & 0.125      & [0.115, 0.134]          \\
        \hline
        $\kappa_{\Phi\beta}$ & 0.123      & [0.115, 0.131]          \\
        $\mu_\Phi$           & 0.0234     & [0.0199, 0.0267]        \\
        \hline
        $s_{\beta N}$        & 2.38       & [2.02, 2.73]            \\
        $\mu_\beta$          & 2.67       & [2.24, 3.10]            \\
        \hline
        $s_{\alpha\Phi}$     & 0.574      & [0.295, 0.786]          \\
        $\mu_\alpha$         & 8.91       & [6.07, 11.1]            \\
        $f_{\beta\alpha}$    & 0.577      & [0.361, 0.738]          \\
        \hline
    \end{tabular}
    \caption[Inferred parameter values of model 5]
    {Inferred parameter values of model 5. MORE DETAIL(SUPER maxt-5)}
    \label{table:estimated5}
\end{table}

When we are trying more populations, an `over-fitting' behaviour is observed at the trajectory of macrophage (FIGURE). In this case, the discrepancy of simulated data and observed data comes from the last 2 data point (72 hpl and 120 hpl).

\begin{figure}[h]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/overfit.png}}
    \end{center}

    \caption[Simulated trajectory from the last population of model 5, with factors applied]{Simulated trajectory from the last population of model 5, with factors applied. MORE DETAIL}
    \label{fig:overfit}
\end{figure}

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/model345cmp.png}}
    \end{center}

Based on the model comparison results, we could conclude that model 5 was the best model among the tested models. Uncertainties were observed in early generations,

From population 10 to population 20 model 4 gains advantage but after population 20 model 5 is preferred. Again, as DISCUSSED in the model comparison of model 1, 2 and 3, we considered the the violently changing model probability is related to the inference ability at certain epsilon threshold values (epsilon) and possibly the insufficient exploration of the parameter space. Given the present result, model 5 is regarded as the best model. Some further experiments tried larger population size (10,000 particles per population) and more populations (36 populations), but the resultant model did not have a significantly better fit of the observed data, although much more computation resources are used.

A inferred parameter posterior distribution of model 5 is shown in FIGURE, the estimated mean values is shown in TABLE. SOME DESCRIPTION.

@@

[why last PCs]

[which parameters are most sensitive]

[compared to credible intervals plot]

[what does that mean]

A quick sensitivity analysis of the dynamical systems models can be quantified using principle component analysis (PCA) \cite{Toni}, where the first PCs corresponds to sloppy parameters while the last PCs corresponds to stiff parameters \cite{sensitivity}. PCA can only provide rough approximated sensitivity behaviour. As models 5 won in the model selection, a further interest is to explore the parameter-to-model sensitivity.

\begin{figure}[H]
    \begin{center}
        \resizebox{0.8\hsize}{!}{\includegraphics{fig/pc_pie_all.png}}
    \end{center}

    \caption[TODO]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc}
\end{figure}

@@

    \caption[PCA results of the last population of model 5]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc_pie}
\end{figure}

FIGURE shows the results of PCA on the last population of inferred model 5. The last PC (PC12) is mostly consists of the linear combination of $a$ (from the equation of neutrophil) and $\mu_\Phi$ (from the equation of macrophage), to which the model is most sensitive. If we conclude the last 5 PCs (FIGURE), then $v_{N\Phi}$, $\mu_N)$ (from the equation of neutrophil) and $k_{\Phi\beta}$ (from the equation of macrophage) contribute most portions together with $a$ and $\mu_\Phi$. As a conclusion, model 5 is sensitive to changes in the above parameters, which all come from the equation of neutrophil and macrophage i.e. equations of the cells; it may suggests that given the observed data, the inferred model is more sensitive to cells' dynamic rather than cytokines' dynamics.

Also, some visualisations of the last population agreed with conclusions from PCA. Approximated posteriors of the last population plotted in FIGURE and credible ranges of parameters across populations in FIGURE shows that the 5 parameters identified by PCA is indeed `stiff' parameters.

>>>>

This subsection aims to preliminarily explore options in ABC SMC implementations. From the above results, we adopted multivariate normal kernel as the kernel for our real inference on observed data; Local M-NN kernels were also planned to be tested. In the real inference, 25:75 factor were also tried to see if it could help in recovering more features and resulting a better fit. because the improvement by using less data was inapparent, we did not try to change the target summary data size. Prior intervals were explored explicitly and gradually shrunk for our models.

@@

% [separated ABC run on model 1, 2 and 3]

@@

When we are trying more generations with factors, an `over-fitting' behaviour was observed at the trajectory of macrophage (Figure \ref{fig:overfit}). In this case, the discrepancy of simulated data and observed data comes from the last 2 data point (72 hpl and 120 hpl). MORE

\begin{figure}[H]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/overfit.png}}
    \end{center}

    \caption[Simulated trajectory from the last population of model 5, with factors applied]{Simulated trajectory from the last population of model 5, with factors applied. MORE DETAIL}
    \label{fig:overfit}

    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/model345cmp.png}}
    \end{center}

Based on the model comparison results, we could conclude that model 5 was the best model among the tested models. Uncertainties were observed in early generations, where similar patterns were also observed in the model comparison of model 1, 2 and 3. We considered the the violently changing model probability in early generations is related to the inference ability at high epsilon threshold values (epsilon) and possibly the insufficient exploration of the parameter space. However as we only focused on the best model with small final threshold $\epsilon_T$, the easiest way is to extended the number of generation until epsilon converges, or increase the population size for a more statistically confident probability.

% From population 10 to population 20 model 4 gains advantage but after population 20 model 5 is preferred. Again, as DISCUSSED in the model comparison of model 1, 2 and 3, Given the present result, model 5 is regarded as the best model.

% Some further experiments tried larger population size (10,000 particles per population) and more populations (36 populations), but the resultant model did not have a significantly better fit of the observed data, although much more computation resources are used.

For the best model (model 5), the inferred parameter posterior distribution of model 5 is shown in Figure \ref{fig:model5_para}, the estimated mean values is shown in Table \ref{table:estimated5}. The (pair-wise) joint density distribution is shown in Figure \ref{fig:model5_para_joint}. SOME DETAIL.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|c c c|}
        \hline
        Parameter            & Mean value & Inter-quartile interval \\[0.5ex]
        \hline\hline
        $\lambda_N$          & 13.5       & [12.9, 14.2]            \\
        $a$                  & 0.0175     & [0.0145, 0.0207]        \\
        $\kappa_{N\beta}$    & 0.176      & [0.109, 0.234]          \\
        $\mu_N$              & 0.0239     & [0.0113, 0.0332]        \\
        $\nu_{N\Phi}$        & 0.125      & [0.115, 0.134]          \\
        \hline
        $\kappa_{\Phi\beta}$ & 0.123      & [0.115, 0.131]          \\
        $\mu_\Phi$           & 0.0234     & [0.0199, 0.0267]        \\
        \hline
        $s_{\beta N}$        & 2.38       & [2.02, 2.73]            \\
        $\mu_\beta$          & 2.67       & [2.24, 3.10]            \\
        \hline
        $s_{\alpha\Phi}$     & 0.574      & [0.295, 0.786]          \\
        $\mu_\alpha$         & 8.91       & [6.07, 11.1]            \\
        $f_{\beta\alpha}$    & 0.577      & [0.361, 0.738]          \\
        \hline
    \end{tabular}
    \caption[Inferred parameter values of model 5]
    {Inferred parameter values of model 5. MORE DETAIL(SUPER maxt-5)}
    \label{table:estimated5}
\end{table}

@@

% [why last PCs]

% [which parameters are most sensitive]

% [compared to credible intervals plot]

% [what does that mean]

A quick sensitivity analysis of the dynamical systems models can be quantified using principle component analysis (PCA) \cite{Toni}, where the first PCs corresponds to sloppy parameters while the last PCs corresponds to stiff parameters \cite{sensitivity}. PCA can only provide rough approximated sensitivity behaviour. By identifying the component loadings, the PCs can be visualised in pie charts and used to indicate the sensitivity.

@@

    \caption[PCA results of the last population of model 5]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc}
\end{figure}

\begin{figure}
    \begin{center}
        \resizebox{0.8\hsize}{!}{\includegraphics{fig/pc_pie_all.png}}
    \end{center}

    \caption[TODO]{PCA results of the last population of model 5. MORE DESCRIPTION}
    \label{fig:pc_pie}
\end{figure}

As models 5 won in the model selection, a further interest was to explore the parameter-to-model sensitivity of the resultant model. PCA was performed on the last generation of inferred model 5, Figure \ref{fig:pc} shows the total explained variance by each PC. The first PC explained a major part of the variance (91.5\%), and the rest PCs explained less less than 5\% of the variance each. PC 5 to PC 12 explained much less variance (less than 0.1\% of the total variance) than the first four PCs, thus they were consider to extend much narrower regions on the posterior distributions and could be used to identify parameter that the model is sensitive to.

Figure \ref{fig:pc_pie} shows the fractions of the PCs explained by individual parameters, and a sum of the fractions in the last 8 PCs, i.e. PC 5 to PC 12. The last PC (PC12) is mostly the linear combination of $a$ (from the equation of neutrophil) and $\mu_\Phi$ (from the equation of macrophage), to which the model is most sensitive. If we conclude the last 8 PCs, it can be found that the biggest portions ($v_{N\Phi}$, $a$, $\kappa_{N\beta}$, $\kappa_{\Phi\beta}$, $\mu_N$, $\mu_\Phi$, $\nu_{N\Phi}$) mostly come from the equation of neutrophil and macrophage i.e. equations of the cells (the un-exploded portion in Figure \ref{fig:pc_pie} a); if we look at the last 8 PCs individually, the same pattern can be observed. This might suggest that the model is more sensitive to cells' dynamic rather than cytokines' dynamics.

The credible intervals (Figure \ref{fig:credible}) of each parameter across generations also agreed with the above conclusion from PCA. The stiff parameters, e.g. $a$ and $\mu_N$, extend narrow credible intervals, while sloppy parameters, e.g. $\mu_\alpha$ and $s_{\alpha\Phi}$, have wider credible intervals across the last few generations.

% then $v_{N\Phi}$, $\mu_N)$ (from the equation of neutrophil) and $k_{\Phi\beta}$ (from the equation of macrophage) contribute most portions together with $a$ and $\mu_\Phi$. As a conclusion, model 5 is sensitive to changes in the above parameters, which all come from the equation of neutrophil and macrophage i.e. equations of the cells; it may suggests that given the observed data, the inferred model is more sensitive to cells' dynamic rather than cytokines' dynamics.

% Also, some visualisations of the last population agreed with conclusions from PCA. Approximated posteriors of the last population plotted in FIGURE and credible ranges of parameters across populations in FIGURE shows that the 5 parameters identified by PCA is indeed `stiff' parameters.

### 2020-08-16 performance/performance.tex
New figures
New fix
New content

    [MORE meanings of the performance study]

@@

[HOW PYABC parallelise the sampling]

@@

[scaling-up performance: speed-up and efficiency]

[large variation in required sampling numbers]

[possible reasons]

The recorded execution time using different number of cores is shown in FIGURE. According to that, speed-up and efficiency can be calculated. Here 8 cores is regarded as the baseline for speed-up and efficiency, as the serial version (using only one core) can take quite a long time to finish 10 repeats. When increasing number of cores, the execution time drops fast at first but the decreasing rate (dropping speed, i.e. the derivative of the execution time against time) is gradually reduced to zero; 36, 54 and 72 cores gives nearly the same average execution time and consequently gains very close speedup values (FIGURE).

@@

A high variance of total required samples was observed in the scaling-up experiments: some single runs required much more samples to finish 20 populations. Moreover, we found that the execution time is not ideally proportional to the total required samples (FIGURE). A possible explain to these phenomenons can be that the median epsilon schedule results in different threshold strategies due to the randomness in the sampling process, thus results in different approximate convergency paths for different runs. Some runs that are stuck in a local optimum and require smaller target epsilon will take huge amount of samples to move away from the local optimum. This have been observed when the distribution of posterior is plotted before and after jumping away from local modes (FIGURE and more explain). Also, time spent in each sample using sanme number of cores is not constant in our case, which might be a result of the parallel schedule; ideally a serial run using only one core will see a nearly constant time spent in every particle.

Due to the unstable sampling numbers, our interests switched to the per-particle performance, where the average sampling numbers per second (FIGURE) and average time spent in one particle (FIGURE) under different of cores ($p$) are plotted. The log-like average time per particle curve shows that using more cores can improve the sampling efficiency, but when $p$ is at a high level (e.g. 36 in our case), the improvement will be less significant.

\section{Profiling}

The performance could also be analysed given a profiling report. The second experiment profiles the program to reveal the detailed time consumption for each operation and the possible bottleneck, according to which we could find the hot-spot of program and given possible suggestions on improving the performance.

In this case, profile tools \verb|cProfile| and \verb|yappi| is used in PyCharm IDE.

[profiling results: hot-spot and possible improvements]

>>>>

% [MORE meanings of the performance study]

@@

% [HOW PYABC parallelise the sampling]

@@

% [scaling-up performance: speed-up and efficiency]

% [large variation in required sampling numbers]

% [possible reasons]

\begin{figure}[h]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/speedup.png}}
    \end{center}

    \caption[TODO]{Speedup. MORE DESCRIPTION}
    \label{fig:speedup}
\end{figure}

The recorded execution time using different number of cores were used to calculate the speedup of the program (Figure \ref{fig:speedup}). Here 8 cores is regarded as the baseline for speed-up and efficiency, as the serial version (using only one core) can take quite a long time to finish 10 repeats. When increasing number of cores, the execution time drops fast at first but the decreasing rate (dropping speed, i.e. the derivative of the execution time against time) is gradually reduced to zero; 36, 54 and 72 cores gives nearly the same average execution time and consequently gains very close speedup values (Table \ref{table:performance}).

\begin{table}[H]
    \centering
    \begin{tabular}{|c c c c c|}
        \hline
        Number of cores & Time (second) & Avg. required samples & Speedup & Efficiency \\ [0.5ex]
        \hline\hline
        8               & 6617          & 445406.8              & 1       & 1          \\
        16              & 3926          & 590221.0              & 1.61    & 0.807      \\
        24              & 2921          & 573339.8              & 2.17    & 0.723      \\
        36              & 2096          & 446948.7              & 3.02    & 0.672      \\
        54              & 2076          & 640404.9              & 3.05    & 0.452      \\
        72              & 2071          & 768096.0              & 3.06    & 0.340      \\
        \hline
    \end{tabular}
    \caption{Performance data}
    \label{table:performance}
\end{table}

@@

\begin{figure}[h]
    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/sample_per_time.png}}
    \end{center}

    \caption[TODO]{Samples per second. MORE DESCRIPTION}
    \label{fig:sample_per_sec}
\end{figure}

A high variance of total required samples was observed in the scaling-up experiments: some single runs required much more samples to finish 20 populations, where they were supposed to have relatively similar average required samples such that we can compare the performance directly via speedup or efficiency.

Due to the variant required samples, our interests switched to the per-second performance, where the average sampling numbers per second under different of number of cores ($p$) are plotted (Figure \ref{fig:sample_per_sec}). The plot shows that even theoretical speedup stops growing when $p>36$, the average sampling numbers per second performance still get improved when using more than 36 cores. It proves that the implementation used here scales reasonably well without encountering bottleneck when using multi-core parallelisation inside one physical node.

The variant total required samples for different runs under the same setting was most likely the result of different threshold evolve path. Median epsilon schedule and randomness in the sampling process resulted in different sequences of threshold for different runs, which means different repeated runs may have different convergency path \cite{threshold}. For example, Some runs temporarily converges to a local optimum, while the target epsilon in the next generation cannot possibly be reached if the new particles are still concentrated

will take huge amount of samples to move away from the local optimum

Moreover, we found that the execution time was not ideally proportional to the sampling tires in that period of time, for both different repeated runs and different generations in a single run. To illustrate this, a single run from $p=24$ was selected and visualised.

A possible explain to these phenomenons can be that the median epsilon schedule results in different threshold strategies due to the randomness in the sampling process, thus results in different approximate convergency paths for different runs. Some runs that are stuck in a local optimum and require smaller target epsilon will take huge amount of samples to move away from the local optimum. This have been observed when the distribution of posterior is plotted before and after jumping away from local modes (FIGURE and more explain). Also, time spent in each sample using sanme number of cores is not constant in our case, which might be a result of the parallel schedule; ideally a serial run using only one core will see a nearly constant time spent in every particle.

% \section{Profiling}

% The performance could also be analysed given a profiling report. The second experiment profiles the program to reveal the detailed time consumption for each operation and the possible bottleneck, according to which we could find the hot-spot of program and given possible suggestions on improving the performance.

% In this case, profile tools \verb|cProfile| and \verb|yappi| is used in PyCharm IDE.

% \subsubsection{Results}

% [profiling results: hot-spot and possible improvements]
